Numerical trajectory patterns of floating macroalgae

### 2020-10-16 thesis_chapter_3.Rmd
added to methods section. basic stats table for morphology added to appendix. general tidy.

---
title: 'Numerical trajectory patterns of floating macroalgae'
author: "Ross Coppin"
output:
  html_document:
  word_document:
    reference_docx: markdown/thesis_template.docx
  pdf_document:
    fig_caption: yes
bibliography: references.bib
csl: markdown/frontiers.csl.txt
---

@@

The region of interest for this study was the Cape Peninsula in the Western Cape, South Africa. This was further sub-divided into two domains, the Atlantic domain and False Bay domain (see figure \ref{fig:map}A). One release site is located in each of the domains, namely Kommetjie in the Atlantic and Buffelsbaai in False Bay (see figure \ref{fig:map}B). The hydrodynamics between the two domains are driven by a complex interaction of wind and hydrodynamic vectors, with wind driven processes primarily occurring in False Bay. For a more detailed explanation of the hydrodynamics of these two domains the reader is referred to @coppin2020. The hydrodynamic characteristics of these two domains are also influenced by larger oceanic processes, which in South Africa can largely be attributed to the Agulhas Current (AC) and the Benguela Current (BC). The AC is a component of the South-west Indian Ocean sub-gyre, flowing predominantly southwestward following the continental shelf edge and eventually retroflecting eastward back into the South Indian Ocean.

@@

The model is based on parameterisations that have been used for previous studies investigating iceberg, capsized marine vessels and microplastics. It should be noted that most of the previous studies regarding macroalgae trajectory have investigated rafts and not solitary floating individuals. A kelp raft will vary in size and can range from a few meters across to large dense mats the size of a sports field. Furthermore, some authors suggest that most macroalgae become entangled through ocean currents and not through the hydrodynamic forces that dislodge them. In addition, large kelp rafts are not a characteristics around the coast of the Cape Peninsula but rather solitary kelp are (Pers. Obs). Therefore, the trajectory of solitary *E. maxima* individuals will be considered. Previous studies have made the assumption that kelp rafts move with the same velocity as the ocean current, and instead model a ensemble of particles which trajectories are determined by a combination of ocean currents and/or the effects of winds on currents. In addition, these studies assume the particle is fully buoyant.

@@

This study simulated two scenarios; one where the plant is fully submerged and only exposed to seawater drag and the other where parts of the plants are partially exposed resulting in a combination of seawater and air form drag. Unlike previous studies, drag based on shape based coefficients was used in the calculations for determining the reduction in velocity as a result of drag. A momentum energy equation was used to calculate the drag forces for the relevant simulation. This approach has been employed in modeling iceberg drift trajectory [@andersson2017; @lichey2001; @eik2009] and was adapted to suit this particular study. The energy momentum equation used to calculate the drag force exerted on the virtual kelp particle was, $$m({d_u/d_t})=F_d + F_a + F_f$$
where $m$ is kelp mass, $d_u$ the virtual kelp particle velocity, $F_d$ the hydrodynamic drag, $F_a$ is the wind drag, and $F_f$ the surface-current flow. The assumption made was that the mass of the kelp did not change for each virtual kelp particle over the course of the simulations. Hydrodynamic drag, wind drag and surface current-flow are two-dimensional vector quantities and were calculated before input into the model.

@@

In order to incorporate hydrodynamic and wind drag, the cross-sectional area of the kelp was calculated first. Known geometric shapes reflecting the relevant plant sections were used to estimate the surface area for various parts of the plant, for details please see image/table ???. The cross-sectional area calculated was site specific and the dimensions needed were garnered from morphology data from a previous study [@coppin2020] and estimated from figure \ref{fig:morphology}.

```{r morphology_wrangling, message=FALSE, warning=FALSE, include=FALSE}
# Investigate morphology data to assist in estimating surface area
# Load software
library(tidyverse)
library(ggpubr)
library(gridExtra)
library(ggplot2)
library(ggord)
library(vegan)
# Read in morphology data
eck_data <- read.csv("data/morph_data/eck_morph_data.csv", sep = ";") %>%
  filter(site!= "Rocky Bank") %>%   # Not using Rocky Bank in this study
  filter(site!= "De Hoop")
eck_data$site <- as.character(eck_data$site)
eck_deep <- eck_data %>% 
  filter(population == "Deep")

eck_long <- gather(eck_deep, morphology, measurement, frond_mass:total_length, factor_key=TRUE)
# Summary stats to make it easier to refer to later, these will not be used any analysis.
eck_stats <- eck_long %>% 
  group_by(site, morphology) %>% 
  summarise(mean = mean(measurement, na.rm = T),
            min = min(measurement, na.rm = T),
            max = max(measurement, na.rm = T),
            range = range(measurement, na.rm = T)[2]-range(measurement, na.rm = T)[1],
            sd = sd(measurement, na.rm = T)) %>% 
  ungroup()
# Make custom positions
positionsEck <- c("Betty's Bay", "Kalk Bay", "A-Frame", 
                  "Miller's Point", "Baboon Rock", "Batsata Rock", 
                  "Buffelsbaai", "Scarborough", 
                  "Soetwater","Kommetjie","Hout Bay", 
                  "Oudekraal", "St Helena Bay")
# Make plots for each morphology feature and grouped by site.
frond_mass <- ggplot(data = eck_data, aes(x = site, y = frond_mass, fill = site)) +
           geom_boxplot(outlier.shape = NA, varwidth = TRUE, notch = TRUE, show.legend = FALSE) +
           stat_compare_means(method = "kruskal.test", label.x = 5) + 
           scale_x_discrete(limits = positionsEck) + 
           ylab("Frond mass (g)") + 
           coord_flip() +
           xlab("") + 
           theme_bw()
primary_length <- ggplot(data = eck_data, aes(x = site, y = primary_length , fill = site)) +
           geom_boxplot(outlier.shape = NA, varwidth = TRUE, notch = TRUE, show.legend = FALSE) +
           stat_compare_means(method = "kruskal.test", label.x = 5) + 
           scale_x_discrete(limits = positionsEck) + 
           ylab("Primary Length  (cm)") + 
           coord_flip() +
           xlab("") + 
           theme_bw()
primary_width <- ggplot(data = eck_data, aes(x = site, y = primary_width , fill = site)) +
           geom_boxplot(outlier.shape = NA, varwidth = TRUE, notch = TRUE, show.legend = FALSE) +
           stat_compare_means(method = "kruskal.test", label.x = 5) + 
           scale_x_discrete(limits = positionsEck) + 
           ylab("Primary Width (cm)") + 
           coord_flip() +
           xlab("") + 
           theme_bw() 
frond_length <- ggplot(data = eck_data, aes(x = site, y = frond_length , fill = site)) +
           geom_boxplot(outlier.shape = NA, varwidth = TRUE, notch = TRUE, show.legend = FALSE) +
           stat_compare_means(method = "kruskal.test", label.x = 5) + 
           scale_x_discrete(limits = positionsEck) + 
           ylab("Frond length (cm)") + 
           coord_flip() +
           xlab("") + 
           theme_bw() 
stipe_mass <- ggplot(data = eck_data, aes(x = site, y = stipe_mass , fill = site)) +
           geom_boxplot(outlier.shape = NA, varwidth = TRUE, notch = TRUE, show.legend = FALSE) +
           stat_compare_means(method = "kruskal.test", label.x = 5) + 
           scale_x_discrete(limits = positionsEck) + 
           ylab("Stipe mass (g)") + 
           coord_flip() +
           xlab("") + 
           theme_bw() 
stipe_length <- ggplot(data = eck_data, aes(x = site, y = stipe_length , fill = site)) +
           geom_boxplot(outlier.shape = NA, varwidth = TRUE, notch = TRUE, show.legend = FALSE) +
           stat_compare_means(method = "kruskal.test", label.x = 5) + 
           scale_x_discrete(limits = positionsEck) + 
           ylab("Stipe length (cm)") + 
           coord_flip() +
           xlab("") + 
           theme_bw() 
stipe_circ <- ggplot(data = eck_data, aes(x = site, y = stipe_circ , fill = site)) +
           geom_boxplot(outlier.shape = NA, varwidth = TRUE, notch = TRUE, show.legend = FALSE) +
           stat_compare_means(method = "kruskal.test", label.x = 5) + 
           scale_x_discrete(limits = positionsEck) + 
           ylab("Stipe circ (cm)") + 
           coord_flip() +
           xlab("") + 
           theme_bw() 
tufts <- ggplot(data = eck_data, aes(x = site, y = tufts , fill = site)) +
           geom_boxplot(outlier.shape = NA, varwidth = TRUE, notch = TRUE, show.legend = FALSE) +
           stat_compare_means(method = "kruskal.test", label.x = 5) + 
           scale_x_discrete(limits = positionsEck) + 
           ylab("No. of tufts") + 
           coord_flip() +
           xlab("") + 
           theme_bw() 
epi_length <- ggplot(data = eck_data, aes(x = site, y = epi_length , fill = site)) +
           geom_boxplot(outlier.shape = NA, varwidth = TRUE, notch = TRUE, show.legend = FALSE) +
           stat_compare_means(method = "kruskal.test", label.x = 5) + 
           scale_x_discrete(limits = positionsEck) + 
           ylab("Epi Length  (cm)") + 
           coord_flip() +
           xlab("") + 
           theme_bw()

total_length <- ggplot(data = eck_data, aes(x = site, y = epi_length, fill = site)) +
           geom_boxplot(outlier.shape = NA, varwidth = TRUE, notch = TRUE, show.legend = FALSE) +
           stat_compare_means(method = "kruskal.test", label.x = 5) + 
           scale_x_discrete(limits = positionsEck) + 
           ylab("Total length  (cm)") + 
           coord_flip() +
           xlab("") + 
           theme_bw()
```

```{r morphology, echo=FALSE, , fig.height=7, fig.width=8, message=FALSE, warning=FALSE, fig.cap='\\label{fig:morphology}Boxplots representing *E.maxima* morphological characteristics by site around the Western Cape coastline, with the *y*-axis depicting the specific morphology measured and with units provided. The lower and upper "hinges" correspond to the first and third quartiles. The whiskers represent the range, solid black lines represent the median and black dots are outliers.'}
# Plot the morphology data based on site (see previous section)
ggarrange(frond_length, frond_mass,primary_length, primary_width, stipe_circ, stipe_length, stipe_mass, total_length, tufts, common.legend = TRUE, legend = c("none"), align = c("hv"))
```

@@

The use of a sphererical shape coefficient was used in the approach to simulate the trajectory of solitary floating *E. maxima*. The drag coefficient for a spherical particle was used in the calculation of hydrodynamic drag as the trajectory model assumes the particles are spherical. The cross-sectional area for each of each virtual plant was converted to the equivalent cross-sectional area for that of a sphere in the equation for calculating hyrdrodynamic drag: $$ F_d = 0.5 \rho A_c C_d U^2$$

@@

Raft-time or sometimes referred to as raft-longevity is the time a macroalgae raft remains afloat at the surface of the ocean after detachment before sinking. Raft-time was estimated from the results of past studies by @graiff2016 and @graiff2013 for each season. For summer the raft time was 25 days, winter was 38 days, and autumn was 41 days respectivley.

@@

## Statisical analysis

@@

>>>>

---
title: 'Numerical trajectory patterns of floating macroalgae'
author: "Ross Coppin"
output:
  html_document:
  pdf_document:
    fig_caption: yes
  word_document:
    reference_docx: markdown/thesis_template.docx
bibliography: references.bib
csl: markdown/frontiers.csl.txt
---

@@

The region of interest for this study was the Cape Peninsula in the Western Cape, South Africa. This was further sub-divided into two domains, the Atlantic domain and False Bay domain (see figure \ref{fig:map}A). One release site is located in each of the domains, namely Kommetjie in the Atlantic and Buffelsbaai in False Bay (see figure \ref{fig:map}B). The hydrodynamics between the two domains are driven by a complex interaction of wind and wave vectors, with wind driven processes primarily occurring in False Bay. For a more detailed explanation of the hydrodynamics of these two domains the reader is referred to @coppin2020. The hydrodynamic characteristics of these regions are also influenced by larger oceanic processes, which in South Africa can largely be attributed to the Agulhas Current (AC) and the Benguela Current (BC). The AC is a component of the South-west Indian Ocean sub-gyre, flowing predominantly southwestward following the continental shelf edge and eventually retroflecting eastward back into the South Indian Ocean.

@@

The model is based on parameterisations that have been used for previous studies investigating iceberg, capsized marine vessels and microplastics. It should be noted that most of the previous studies regarding macroalgae trajectory have investigated rafts and not solitary floating individuals. A kelp raft will vary in size and can range from a few meters across to large dense mats the size of a sports field. Furthermore, some authors suggest that most macroalgae become entangled through ocean currents and not through the hydrodynamic forces that dislodge them. In addition, large kelp rafts are not a characteristics around the coast of the Cape Peninsula but rather solitary kelp are (Pers. Obs). Therefore, the trajectory of solitary *E. maxima* individuals will be considered. Previous studies have made the assumption that floating macroalage move with the same velocity as the ocean current, and instead model a ensemble of particles which trajectories are determined by a combination of ocean currents and/or the effects of winds on currents.

@@

This study simulated two scenarios; one where the plant is fully submerged and only exposed to seawater drag; and the other where parts of the plants are partially exposed resulting in a combination of seawater and air form drag. Drag based on shape based coefficients were used in the calculations for determining the overall kelp velocity vectors. A momentum energy equation was used to calculate the drag forces for the relevant simulation. This approach has been employed in modeling iceberg drift trajectory [@andersson2017; @lichey2001; @eik2009] and was adapted to suit this particular study. The energy momentum equation used to calculate the drag force exerted on the virtual kelp particle was, $$m({d_u/d_t})=F_d + F_a + F_f$$
where $m$ is kelp mass, $d_u$ the virtual kelp particle velocity, $F_d$ the hydrodynamic drag, $F_a$ is the wind drag, and $F_f$ the surface-current flow. The assumption made was that the mass of the kelp did not change for each virtual kelp particle over the course of the simulations. Hydrodynamic drag, wind drag and surface current-flow are two-dimensional vector quantities and were calculatedfor each time step of each simulation.

@@

In order to incorporate hydrodynamic and wind drag, the cross-sectional area of the kelp was calculated first. Known geometric shapes reflecting the relevant plant sections were used to estimate the surface area for various parts of the plant, for details please see image/table ???. The dimensional data needed was estimated in cases were data was not available for that particular morphological characteristic. The bulb/pneumatocyst is a highly variable morphological characteristic and in some cases can appear absent, the same is true for the holdfast. Therefore, the cross-sectional area of the bulb/pneumatocyst was not considered in the calculation of overall cross-sectional area. In terms of the holdfast, a standard cross-sectional area was used for simulations where the inclusion of the holdfast was considered.

The cross-sectional area calculated was site specific and the dimensions needed were garnered from morphology data from a previous study [@coppin2020]. The minimum, maximum and mean were calculated for each morphological characteristic (see appendix), which were used for calculating the overall minimum, maximum and mean cross-sectional areas needed to run the various simulations. The trajectory of these "types" of cross-sectional areas (minimum, maximum and mean overall cross-sectional area) were used to determine the influence of drag, both hydrodynamic and wind, on overall trajectory.

@@

The use of a spherical shape coefficient was used in the approach to simulate the trajectory of solitary floating *E. maxima*. The drag coefficient for a spherical particle was used in the calculation of hydrodynamic drag as the trajectory model assumes the particles are spherical. The cross-sectional area for each of each virtual plant was converted to the equivalent cross-sectional area for that of a sphere in the equation for calculating hydrodynamic drag: $$ F_d = 0.5 \rho A_c C_d U^2$$

@@

Raft-time or sometimes referred to as raft-longevity is the time a macroalgae raft remains afloat at the surface of the ocean after detachment before sinking. Raft-time was estimated from the results of past studies by @graiff2016 and @graiff2013 for each season. For summer the raft time was 25 days, winter was 38 days, and autumn was 41 days respectively.

@@

## Statistical analysis

@@

```{r morphology_wrangling, message=FALSE, warning=FALSE, include=FALSE}
# Investigate morphology data to assist in estimating surface area

# Load software
library(tidyverse)
library(ggpubr)
library(gridExtra)
library(ggplot2)
library(ggord)
library(vegan)
library(dplyr)

# Read in morphology data
eck_df <- read.csv("data/morph_data/eck_morph_data.csv", sep = ";") %>%
  filter(site!= "Rocky Bank") %>%   # Not using Rocky Bank in this study
  filter(site!= "De Hoop")

eck_df$site <- as.character(eck_df$site)
eck_df <- eck_df %>% 
  filter(population == "Deep")

eck_long_df <- gather(eck_df, Morphology, measurement, frond_mass:total_length, factor_key=TRUE)
# Summary stats to make it easier to refer to later, these will not be used any analysis.

site_list <- c("Kommetjie")

eck_stats <- eck_long_df %>% 
  group_by(site, Morphology) %>% 
  summarise(mean = mean(measurement, na.rm = T),
            minimum = min(measurement, na.rm = T),
            maximum = max(measurement, na.rm = T)) %>% 
  ungroup() %>% 
  filter(site %in% site_list) %>% 
  rename(Site = site) %>% 
  rename(Mean = mean) %>% 
  rename(Minimum = minimum) %>% 
  rename(Maximum = maximum)

#df <- eck_stats
#rownames(df)[rownames(df) == "frond_mass"] <- "Frond mass"
#
#  eck_stats <- eck_stats %>% 
#  column_to_rownames(var = "Morphology") %>% 
#  mutate(Morphology = replace(Morphology, 
#                              Morphology == "epi_length", "Epiphyte length (cm)")) %>% 
#  mutate(Morphology = replace(Morphology, 
#                              Morphology == "frond_length", "Frond length (cm)")) %>% 
#  mutate(Morphology = replace(Morphology, 
#                              Morphology == "frond_mass", "Frond mass (kg)")) %>% 
#  mutate(Morphology = replace(Morphology, 
#                              Morphology == "primary_length", "Primary length (cm)")) %>% 
#  mutate(Morphology = replace(Morphology, 
#                              Morphology == "primary_width", "Primary width (cm)")) %>% 
#  mutate(Morphology = replace(Morphology, 
#                              Morphology == "stipe_circ", "Stipe circumference (mm)")) %>% 
#  mutate(Morphology = replace(Morphology, 
#                              Morphology == "stipe_length", "Stipe length (cm)")) %>% 
#  mutate(Morphology = replace(Morphology, 
#                              Morphology == "stipe_mass", "Stipe mass (kg)")) %>% 
#  mutate(Morphology = replace(Morphology, 
#                              Morphology == "total_length", "Total length (cm)")) %>% 
#  mutate(Morphology = replace(Morphology, 
#                              Morphology == "tufts", "Number of tufts"))

```{r echo=FALSE, message=FALSE, warning=FALSE}

library(knitr)
library(kableExtra)

knitr::kable(eck_stats, format = "html") %>%
  kable_styling()
